---
title: "Final_project_script"
output: html_document
date: "2026-04-16"
---

# ----------LOAD THE PACKAGES FOR PROJECT------------
```{r}
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(lubridate)
library(here)
library(vip)          
library(xgboost)      
library(ranger)       
library(tidyverse)    
library(doParallel)   
library(shapviz)
library(tidymodels)
library(recipes)
library(rsample)
library(daymetr)
library(slider)
library(kknn)

set.seed(19823)

```

# ----------LOAD TRAINING DATA------------
```{r}
#| message: false
#| warning: false

train_meta <- read_csv("../data/training/training_meta.csv", show_col_types = FALSE) %>% 
  clean_names()

train_trait <- read_csv("../data/training/training_trait.csv", show_col_types = FALSE) %>% 
  clean_names()

train_soil <- read_csv("../data/training/training_soil.csv", show_col_types = FALSE) %>% 
  clean_names()

```

# ------------AGGREGATE TRAINING TRAITS-------------
```{r}
#| message: false
#| warning: false

train_trait_clean <- train_trait %>%
  mutate(
    date_planted = mdy(date_planted),
    date_harvested = mdy(date_harvested)
  ) %>%
  group_by(site, year, hybrid) %>%
  summarize(
    yield = mean(yield_mg_ha, na.rm = TRUE),
    grain_moisture = mean(grain_moisture, na.rm = TRUE),
    date_planted = min(date_planted, na.rm = TRUE),
    date_harvested = max(date_harvested, na.rm = TRUE),
    .groups = "drop"
  )

head(train_trait_clean)
```

# -------------MERGE------------
```{r}
#| message: false
#| warning: false

train_full <- train_trait_clean %>%
  left_join(train_meta, by = c("site", "year")) %>%
  left_join(train_soil, by = c("site", "year")) %>%
  mutate(
    across(
      any_of(c("soilp_h", "om_pct", "soilk_ppm", "soilp_ppm")),
      as.numeric
    )
  )

glimpse(train_full)

```

# ----------LOAD TESTING DATA------------
```{r}
#| message: false
#| warning: false

test_meta <- read_csv("../data/testing/testing_meta.csv") %>%
  clean_names() %>%
  mutate(
    previous_crop = str_to_lower(previous_crop),
    year = as.integer(year)
  )

test_sub <- read_csv("../data/testing/testing_submission.csv") %>%
  clean_names() %>%
  rename(yield = yield_mg_ha) %>%
  mutate(year = as.integer(year))

head(test_sub)
head(test_meta)

```

# ----------FIXING SOIL SITE COLUMN---------------
```{r}
#| message: false
#| warning: false

test_soil <- read_csv("../data/testing/testing_soil.csv", show_col_types = FALSE) %>%
  clean_names() %>%
  separate(site, into = c("site", "year_str"), sep = "_") %>%
  mutate(
    year = as.integer(year_str)
  ) %>%
  select(-year_str) %>%
  group_by(site, year) %>%
  summarize(
    soilp_h = mean(soilp_h, na.rm = TRUE),
    om_pct = mean(om_pct, na.rm = TRUE),
    soilk_ppm = mean(soilk_ppm, na.rm = TRUE),
    soilp_ppm = mean(soilp_ppm, na.rm = TRUE),
    .groups = "drop"
  )

head(test_soil)
```

# -------------MERGE------------
```{r}
#| message: false
#| warning: false

test_full <- test_sub %>%
  left_join(test_meta, by = c("site", "year")) %>%
  left_join(test_soil, by = c("site", "year"))

glimpse(test_full)

nrow(test_sub)
nrow(test_full)
```

#----------CREATE PLANTING AND HARVEST WINDOWS (BY SITE)------
```{r}
#| message: false
#| warning: false

site_calendar <- train_full %>%
  mutate(
    plant_doy = lubridate::yday(date_planted),
    harvest_doy = lubridate::yday(date_harvested)
  ) %>%
  group_by(site) %>%
  summarize(
    typical_plant_doy = round(median(plant_doy, na.rm = TRUE)),
    typical_harvest_doy = round(median(harvest_doy, na.rm = TRUE)),
    .groups = "drop"
  )

global_plant_doy <- round(median(site_calendar$typical_plant_doy, na.rm = TRUE))
global_harvest_doy <- round(median(site_calendar$typical_harvest_doy, na.rm = TRUE))

head(site_calendar)
names(site_calendar)
```

```{r}
test_full <- test_full %>%
  left_join(site_calendar, by = "site")

names(test_full)
```


#-----------ADDING P/H WINDOWS TO TESTING AND TRAINING DATA------------
```{r}
#| message: false
#| warning: false

train_full <- train_full %>%
  select(-any_of(c(
    "plant_doy", "harvest_doy",
    "typical_plant_doy", "typical_harvest_doy",
    "typical_plant_doy.x", "typical_harvest_doy.x",
    "typical_plant_doy.y", "typical_harvest_doy.y"
  ))) %>%
  mutate(
    plant_doy = lubridate::yday(date_planted),
    harvest_doy = lubridate::yday(date_harvested)
  ) %>%
  left_join(site_calendar, by = "site") %>%
  mutate(
    plant_doy = coalesce(plant_doy, typical_plant_doy, global_plant_doy),
    harvest_doy = coalesce(harvest_doy, typical_harvest_doy, global_harvest_doy)
  )

test_full <- test_full %>%
  select(-any_of(c(
    "plant_doy", "harvest_doy",
    "typical_plant_doy", "typical_harvest_doy",
    "typical_plant_doy.x", "typical_harvest_doy.x",
    "typical_plant_doy.y", "typical_harvest_doy.y"
  ))) %>%
  left_join(site_calendar, by = "site") %>%
  mutate(
    plant_doy = coalesce(typical_plant_doy, global_plant_doy),
    harvest_doy = coalesce(typical_harvest_doy, global_harvest_doy)
  )

grep("\\.x$|\\.y$", names(train_full), value = TRUE)
grep("\\.x$|\\.y$", names(test_full), value = TRUE)

glimpse(test_full)
```
#------CHECK P/H VALUES-------
```{r}
#| message: false
#| warning: false

summary(train_full$plant_doy)
summary(train_full$harvest_doy)

summary(test_full$plant_doy)
summary(test_full$harvest_doy)
```

#-------CREATING A SITE-YEAR LIST-------
```{r}
#| message: false
#| warning: false

all_site_years <- bind_rows(
  train_full %>% distinct(site, year, latitude, longitude, plant_doy, harvest_doy),
  test_full %>% distinct(site, year, latitude, longitude, plant_doy, harvest_doy)
) %>%
  arrange(site, year) %>%
  mutate(
    longitude = if_else(longitude > 0 & longitude < 180, -longitude, longitude)
  )

head(all_site_years)
```

#----------FIXING COORDINATES---------
```{r}
site_coords <- all_site_years %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  group_by(site) %>%
  summarize(
    latitude_fill = first(latitude),
    longitude_fill = first(longitude),
    .groups = "drop"
  )

all_site_years <- all_site_years %>%
  left_join(site_coords, by = "site") %>%
  mutate(
    latitude = coalesce(latitude, latitude_fill),
    longitude = coalesce(longitude, longitude_fill)
  ) %>%
  select(-latitude_fill, -longitude_fill)

summary(all_site_years$plant_doy)
summary(all_site_years$harvest_doy)

# remove site-years with no coordinates available
all_site_years <- all_site_years %>%
  filter(!is.na(latitude), !is.na(longitude))

# final check
all_site_years %>%
  filter(is.na(latitude) | is.na(longitude))
```

#----------PULLING DAYMET DATA---------- -->
```{r}
#| message: false
#| warning: false

# get_daymet_one <- function(site, year, lat, lon) {
#   tryCatch({
#     out <- download_daymet(
#       site = paste0(site, "_", year),
#       lat = lat,
#       lon = lon,
#       start = year,
#       end = year,
#       internal = TRUE,
#       silent = TRUE
#     )
# 
#     out$data %>%
#       as_tibble() %>%
#       mutate(
#         site = site,
#         year = year
#       )
#   }, error = function(e) {
#     message(paste("Skipping:", site, year, "lat =", lat, "lon =", lon))
#     return(NULL)
#   })
# }
# 
# weather_list <- purrr::pmap(
#   all_site_years,
#   ~ get_daymet_one(site = ..1, year = ..2, lat = ..3, lon = ..4)
# )
# 
# weather_daily <- bind_rows(weather_list)
# 
# 
# write_csv(weather_daily,
#           "../data/weather_daily.csv"
#           ) #saving pulled weather data to avoid repeated pulling
```

#------------FEATURE ENGINEERING WEATHER VARIABLES------------
```{r}
#| message: false
#| warning: false

weather_daily <- read_csv("../data/weather_daily.csv", show_col_types = FALSE) %>%
  clean_names() %>%
  mutate(
    mean_temp = (tmax_deg_c + tmin_deg_c) / 2,
    tmax_cap = pmin(tmax_deg_c, 30),
    tmin_cap = pmax(tmin_deg_c, 10),
    gdd = pmax(((tmax_cap + tmin_cap) / 2) - 10, 0),
    hot_days = as.integer(tmax_deg_c > 32),
    hot_nights = as.integer(tmin_deg_c > 21),
    rain_days = as.integer(prcp_mm_day >= 1),
    total_radiation = (srad_w_m_2 * dayl_s) / 1e6
  )

glimpse(weather_daily)
```

#-----------SUMMARIZE WEATHER VARIABLES------------
```{r}
#| message: false
#| warning: false

weather_features <- weather_daily %>%
  group_by(site, year) %>%
  summarize(
    mean_temp = mean(mean_temp, na.rm = TRUE),
    max_temp = max(tmax_deg_c, na.rm = TRUE),
    min_temp = min(tmin_deg_c, na.rm = TRUE),
    total_prcp = sum(prcp_mm_day, na.rm = TRUE),
    rain_days = sum(rain_days, na.rm = TRUE),
    hot_days = sum(hot_days, na.rm = TRUE),
    hot_nights = sum(hot_nights, na.rm = TRUE),
    gdd = sum(gdd, na.rm = TRUE),
    total_radiation = sum(total_radiation, na.rm = TRUE),
    mean_vp = mean(vp_pa, na.rm = TRUE),
    .groups = "drop"
  )

glimpse(weather_features)
```

#-------ADDING WEATHER TO TRAINING AND TESTING DATA------- -->
```{r}
#| message: false
#| warning: false

weather_cols <- c(
  "mean_temp", "max_temp", "min_temp", "total_prcp", "rain_days",
  "hot_days", "hot_nights", "gdd", "total_radiation", "mean_vp"
)

train_full <- train_full %>%
  select(-any_of(c(weather_cols, paste0(weather_cols, ".x"), paste0(weather_cols, ".y")))) %>%
  left_join(weather_features, by = c("site", "year"))

test_full <- test_full %>%
  select(-any_of(c(weather_cols, paste0(weather_cols, ".x"), paste0(weather_cols, ".y")))) %>%
  left_join(weather_features, by = c("site", "year"))

# sanity check
grep("\\.x$|\\.y$", names(train_full), value = TRUE)
grep("\\.x$|\\.y$", names(test_full), value = TRUE)

glimpse(train_full)
glimpse(test_full)
```

#-----------SAVING THE FINISHED AND CLEAN DATA--------- -->
```{r}
#| message: false
#| warning: false

nrow(train_full)
nrow(test_full)

grep("\\.x$|\\.y$", names(train_full), value = TRUE)
grep("\\.x$|\\.y$", names(test_full), value = TRUE)

write_csv(train_full, "../data/train_full_weather.csv")
write_csv(test_full, "../data/test_full_weather.csv")
```

#--------LOADING FINISHED DATA SETS-------
```{r}
#| message: false
#| warning: false

train_full <- read_csv("../data/train_full_weather.csv", show_col_types = FALSE) %>%
  mutate(
    date_planted = lubridate::ymd(date_planted),
    date_harvested = lubridate::ymd(date_harvested)
  )

test_full  <- read_csv("../data/test_full_weather.csv", show_col_types = FALSE)

glimpse(train_full)
glimpse(test_full)
```

```{r}
numeric_cols <- c(
  "soilp_h", "om_pct", "soilk_ppm", "soilp_ppm",
  "plant_doy", "harvest_doy",
  "typical_plant_doy", "typical_harvest_doy",
  "mean_temp", "max_temp", "min_temp", "total_prcp", "rain_days",
  "hot_days", "hot_nights", "gdd", "total_radiation", "mean_vp"
)

clean_numeric <- function(x) {
  if (is.character(x)) {
    readr::parse_number(x)
  } else {
    as.numeric(x)
  }
}

train_full <- train_full %>%
  mutate(across(any_of(numeric_cols), clean_numeric))

test_full <- test_full %>%
  mutate(across(any_of(numeric_cols), clean_numeric))
```


#-------USING 2023 AS A TEST SET-------
```{r}
#| message: false
#| warning: false

set.seed(23628)

train_dev <- train_full %>%
  filter(year < 2023) %>%
  select(-grain_moisture, -date_planted, -date_harvested) %>%
  mutate(
    site = as.character(site),
    hybrid = as.character(hybrid),
    previous_crop = as.character(previous_crop)
  )

test_2023 <- train_full %>%
  filter(year == 2023) %>%
  select(-grain_moisture, -date_planted, -date_harvested) %>%
  mutate(
    site = as.character(site),
    hybrid = as.character(hybrid),
    previous_crop = as.character(previous_crop)
  )

train_dev_knn <- train_dev %>%
  slice_sample(n = min(20000, nrow(train_dev)))

nrow(train_dev)
nrow(test_2023)
nrow(train_dev_knn)
```

#---------CREATING THE RECIPES---------
```{r}
#| message: false
#| warning: false

corn_recipe_xgb <- recipe(yield ~ ., data = train_dev) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors())

knn_recipe <- recipe(yield ~ ., data = train_dev) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())
```

#---------FITTING XGBOOST---------
```{r}
#| message: false
#| warning: false

xgb_spec <- boost_tree(
  trees = 800,
  tree_depth = 6,
  learn_rate = 0.03,
  min_n = 5,
  loss_reduction = 0,
  sample_size = 0.8,
  mtry = 10
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

xgb_wf <- workflow() %>%
  add_recipe(corn_recipe_xgb) %>%
  add_model(xgb_spec)

xgb_fit <- fit(xgb_wf, data = train_dev)

xgb_pred_2023 <- predict(xgb_fit, new_data = test_2023) %>%
  bind_cols(test_2023 %>% select(site, year, hybrid, yield))

xgb_metrics <- metrics(xgb_pred_2023, truth = yield, estimate = .pred)
xgb_metrics
```

#---------FITTING K-NEAREST NEIGHBORS---------
```{r}

knn_spec <- nearest_neighbor(
  neighbors = 5,
  weight_func = "rectangular",
  dist_power = 2
) %>%
  set_engine("kknn") %>%
  set_mode("regression")

knn_wf <- workflow() %>%
  add_recipe(knn_recipe) %>%
  add_model(knn_spec)

knn_fit <- fit(knn_wf, data = train_dev_knn)

knn_pred_2023 <- predict(knn_fit, new_data = test_2023) %>%
  bind_cols(test_2023 %>% select(site, year, hybrid, yield))

knn_metrics <- metrics(knn_pred_2023, truth = yield, estimate = .pred)
knn_metrics
```

#------------PLOTTING XGBOOST------------
```{r}
xgb_rmse <- xgb_metrics %>% 
  filter(.metric == "rmse") %>% 
  pull(.estimate)
xgb_rsq  <- xgb_metrics %>% 
  filter(.metric == "rsq") %>% 
  pull(.estimate)

ggplot(xgb_pred_2023, aes(x = yield, y = .pred)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "XGBoost: Predicted vs Observed Yield",
    subtitle = paste0("RMSE = ", round(xgb_rmse, 2), " | R² = ", round(xgb_rsq, 3)),
    x = "Observed Yield",
    y = "Predicted Yield"
  ) +
  theme_minimal()
```

#------------PLOTTING K-NEAREST NEIGHBOR-----------
```{r}
knn_rmse <- knn_metrics %>% 
  filter(.metric == "rmse") %>% 
  pull(.estimate)
knn_rsq  <- knn_metrics %>% 
  filter(.metric == "rsq") %>% 
  pull(.estimate)

ggplot(knn_pred_2023, aes(x = yield, y = .pred)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "kNN: Predicted vs Observed Yield",
    subtitle = paste0("RMSE = ", round(knn_rmse, 2), " | R² = ", round(knn_rsq, 3)),
    x = "Observed Yield",
    y = "Predicted Yield"
  ) +
  theme_minimal()
```

#--------------COMPARING MODELS------------
```{r}
#| message: false
#| warning: false
model_comparison <- bind_rows(
  xgb_metrics %>% 
    mutate(model = "XGBoost"),
  knn_metrics %>% 
    mutate(model = "K-Nearest Neighbor (kNN)")
)

print(model_comparison)
```

#-----------REFIT FINAL MODEL ON ALL TRAINING DATA--------
```{r}
#| message: false
#| warning: false

final_train <- train_full %>%
  select(-grain_moisture, -date_planted, -date_harvested) %>%
  mutate(
    site = as.character(site),
    hybrid = as.character(hybrid),
    previous_crop = as.character(previous_crop)
  )

final_xgb_fit <- fit(xgb_wf, data = final_train)
#final_knn_fit <- fit(knn_wf, data = final_train)
```

#------------PREDICT 2024-------------------
```{r}
#| message: false
#| warning: false

final_test <- test_full %>%
  select(-yield) %>%
  mutate(
    site = as.character(site),
    hybrid = as.character(hybrid),
    previous_crop = as.character(previous_crop)
  )

xgb_pred_2024 <- predict(final_xgb_fit, new_data = final_test) %>%
  bind_cols(final_test %>% select(site, year, hybrid)) %>%
  rename(predicted_yield = .pred)

#knn_pred_2024 <- predict(final_knn_fit, new_data = final_test) %>%
  bind_cols(final_test %>% select(site, year, hybrid)) %>%
  rename(predicted_yield = .pred)

head(xgb_pred_2024)
#head(knn_pred_2024)
```

#--------------SAVE PREDICTIONS------------
```{r}
write_csv(xgb_pred_2024, "../output/xgb_predictions_2024.csv")
write_csv(knn_pred_2024, "../output/knn_predictions_2024.csv")
```

#-------------ADDING TO SUBMISSION FILE-------------
```{r}
submission <- read_csv("../data/testing/testing_submission.csv", show_col_types = FALSE)

submission_xgb <- submission %>%
  mutate(yield_mg_ha = xgb_pred_2024$predicted_yield)

submission_knn <- submission %>%
  mutate(yield_mg_ha = knn_pred_2024$predicted_yield)

write_csv(submission_xgb, "../output/final_submission_xgb.csv")
write_csv(submission_knn, "../output/final_submission_knn.csv")

head(submission_xgb)
```


#-------------VARIABLE IMPORTANCE------------
```{r}
final_xgb_fit %>%
  extract_fit_parsnip() %>%
  vip(num_features = 15)
```


```{r}
knitr::purl("finalproject_code.qmd", output = "finalproject_script.R", documentation = 0)
```

