---
title: "Final Proejct"
format: html
editor: visual
---

# ----------LOAD TRAINING DATA------------
```{r}
#| message: false
#| warning: false

train_meta <- read_csv("../data/training/training_meta.csv") %>% 
  janitor::clean_names() 
train_trait <- read_csv("../data/training/training_trait.csv") %>% 
  janitor::clean_names() 
train_soil <- read_csv("../data/training/training_soil.csv") %>% 
  janitor::clean_names()


train_full <- train_trait %>%
  left_join(train_meta, by = c("site", "year"))


#correct some coordinates
train_full <- train_full %>% 
  mutate(
    longitude = if_else(longitude > 0 & longitude < 180, -longitude, longitude)
  )
```


```{r}
# #-----------------DOWNLOAD WEATHER DATA------------------
# 
daymet_all <- train_full %>%
    mutate(weather = pmap(list(year, site, latitude, longitude),
                        function(.y, .site, .lat, .lon) {
                          tryCatch({
                            daymetr::download_daymet(
                              site = paste0(.site, "_", .y),
                              lat = .lat,
                              lon = .lon,
                              start = .y,
                              end = .y,
                              internal = T,
                              silent = T,
                              simplify = T)$data %>%
                              as_tibble()}, error = function(e) NULL) }
                        ))

beepr::beep()
```


```{r}
#---------------------UNNEST AND SAVE WEATHER DATA-------------------

daymet_all_unnes <- daymet_all %>%
  unnest(weather) %>%
  pivot_wider(names_from = measurement,
              values_from = value
              ) %>%
    janitor::clean_names()

head(daymet_all_unnes)

daymet_all_unnest <- daymet_all_unnes %>% #full unnesting
  unnest(dayl_s:vp_pa)

head(daymet_all_unnest)
# Exporting

write_csv(daymet_all_unnest,
          "../data/weatherdata.csv"
          )
```


#------CONTINUE LOADING DATA--------------
```{r}
#| message: false
#| warning: false

train_weather <- read_csv("../data/weatherdata.csv") %>% 
  clean_names()
```


#-----------CLEAN META--------------
```{r}
#| message: false
#| warning: false
 train_meta <- train_meta %>%
 mutate(previous_crop = str_to_lower(previous_crop)) 
```



# -------------MERGE TRAINING DATA------------
```{r}
#| message: false
#| warning: false

train_full <- train_weather %>%
    left_join(train_soil, by = c("site", "year")) %>% 




head(train_full)
summary(train_full$year)
table(train_full$site)
```

#-------------GROUP BY SITE_YEAR AND CONVERT PREDICTORS AS FACTORS---------------------
```{r }
#| message: false
#| warning: false

train_full <- train_full %>%
  mutate(site_year = paste(site, year, sep = "_")) %>%
  mutate(
    site_year     = factor(site_year),
    hybrid        = factor(hybrid),
    previous_crop = factor(previous_crop),
    site          = factor(site)
  )
```


```{r }
#| message: false
#| warning: false

parse_mdy_year <- function(x, year) {
    suppressWarnings(
    lubridate::mdy(paste(x, year))
  )
}

# count of predictor columns BEFORE recipe steps
n_pred <- train_full %>%
  select(-yield_mg_ha, -site_year) %>%
  ncol()

n_pred
```

#--------NOW LET US DEFINE THE RECIPE TO USE-----

```{r}
#| message: false
#| warning: false

soybean_recipe <- recipe(
  yield_mg_ha ~ year + site + replicate + block + hybrid + date_planted + date_harvested + 
    grain_moisture + previous_crop + longitude + latitude + soilp_h + om_pct + soilk_ppm + soilp_ppm,  data = train_full) %>%
  
  step_string2factor(all_nominal_predictors()) %>%
  
  # Create Date variables from the character date columns
  step_mutate(
    planted_date   = mdy(date_planted),
    harvested_date = mdy(date_harvested)
  ) %>%
  
  # Engineer numeric date features
  step_mutate(
    doy_planted      = yday(planted_date),
    doy_harvested    = yday(harvested_date),
    harvest_interval = as.numeric(harvested_date - planted_date)
  ) %>%
  
  # Remove original string and intermediate Date columns
  step_rm(date_planted, date_harvested, planted_date, harvested_date) %>%
  
  # Impute missing numeric and categorical data
  step_impute_median(all_numeric_predictors()) %>%   # numeric median imputation[web:37][web:43]
  step_impute_mode(all_nominal_predictors()) %>%     # mode imputation for factors[web:40]
  
  # Dummy encode categorical predictors
  step_dummy(all_nominal_predictors())

soybean_recipe
```

#---------------RESAMPLING GROUP---------------
```{r}
#| message: false
#| message: false

set.seed(5253626)

cv_folds <- group_vfold_cv(
  train_full,
  group = site_year,
  v = 10
)

soybean_recipe
cv_folds

```

#----------------------LET'S TUNE THE MODEL XGBOOST HYPERPARAMETERS-----------------------

```{r}
#| message: false
#| message: false


xgboost_m <- boost_tree(
  trees          = 1000,
  tree_depth     = tune(),
  learn_rate     = tune(),
  loss_reduction = tune(),
  min_n          = tune(),
  sample_size    = tune(),
  mtry           = tune()
) %>%
  set_mode("regression") %>%
  set_engine("xgboost")

xgboost_m
```


#---------------- THE MODEL WORKFLOW------------------------------

```{r}
#| message: false
#| message: false

xgboost_wrkflow <- workflow() %>%
  add_model(xgboost_m) %>%
  add_recipe(soybean_recipe)

```

## ----------------GRID TUNE-----------------------

```{r}
#| message: false
#| message: false

set.seed(5253626)

xgboost_grid_tune <- grid_space_filling(
  tree_depth(),
  learn_rate(range = c(-3, -0.5)),  
  loss_reduction(),
  min_n(),
  sample_size = sample_prop(),
  mtry(range = c(6, min(50, n_pred))),
  size = 30
)

```

#---------------- CV TUNNING--------------

```{r}
#| message: false
#| message: false

set.seed(5253626)

xgboost_tune <- xgboost_grid_tune(
  xgboost_wrkflow,
  resamples = cv_folds,
  grid      = xgboost_grid_tune,
  metrics   = metric_set(rmse, rsq),
  control   = control_grid(save_pred = TRUE)
)

collect_metrics(xgboost_tune) %>%
  filter(.metric == "rmse") %>%
  arrange(mean)

```

#-------------CHECKING BEST XGBOOST PARAMETERS-------------------

```{r}
best_xgboost <- select_best(xgboost_tune, metric = "rmse")

# Finalize the workflow
xgboost_final_wrkflow <- finalize_workflow(xgb_wf, best_xgb)
```

#----------PREDICTING 2024--------------

```{r}
xgboost_final_fit <- fit(xgboost_final_wrkflow, data = train_full)

xgboost_predict <- predict(xgboost_final_fit, new_data = test_full) %>%
  bind_cols(test_full %>% select(year, site, hybrid)) %>%
  rename(pred_yield_mg_ha = .pred)

xgboost_predict
```





## **Specify a random forest model**

```{r}
# rf_spec <- rand_forest(
#   mtry  = tune(),   # number of predictors tried at each split (we'll tune)
#   trees = 500,      # reasonable default number of trees
#   min_n = tune()    # minimum data points in a terminal node
# ) %>%
#   set_mode("regression") %>%
#   set_engine("ranger", importance = "impurity")


rf_spec <- rand_forest(
  mtry  = 20,     # heuristic; can adjust later
  trees = 500,
  min_n = 5
) %>%
  set_mode("regression") %>%
  set_engine("ranger", importance = "impurity")
```

## **Build a workflow (recipe + model)**

```{r}
corn_wf <- workflow() %>%
  add_model(rf_spec) %>%
  add_recipe(corn_rec)

corn_wf
```

## **Evaluate with grouped CV (no tuning)**

```{r}
set.seed(123)

rf_res <- fit_resamples(
  corn_wf,
  resamples = cv_folds,
  metrics   = metric_set(rmse, rsq),
  control   = control_resamples(save_pred = TRUE)
)

collect_metrics(rf_res)
```

## **Fit final model on all training data and predict on test**

```{r}
final_fit <- fit(corn_wf, data = train_df)
```

# Generate predictions for 2024

```{r}
test_pred <- predict(final_fit, new_data = test_df) %>%
  bind_cols(test_df %>% select(year, site, hybrid))

# Rename prediction column
test_pred <- test_pred %>%
  rename(pred_yield_mg_ha = .pred)

test_pred

write_csv(test_pred, "../output/corn_yield_predictions_rf.csv")
```

## **Set up tuning grid and metrics**

```{r}
# metrics
corn_metrics <- metric_set(rmse, rsq)  # yardstick metrics for numeric outcomes[web:93][web:90]

# a small tuning grid
rf_grid <- grid_regular(
  mtry(range = c(5L, 50L)),   # you can widen later
  min_n(range = c(5L, 25L)),
  levels = 5                  # 5x5 = 25 combos
)
```

## **Tune with grouped CV**

```{r}
# training step
set.seed(123)

rf_tuned <- tune_grid(
  corn_wf,
  resamples = cv_folds,
  grid      = rf_grid,
  metrics   = corn_metrics,
  control   = control_grid(save_pred = TRUE)
)

rf_tuned
```

# inspect results

```{r}
collect_metrics(rf_tuned) %>%
  arrange(mean)
```

## **best parameters and finalize**

```{r}
# picks the parameter combination with lowest RMSE over the resamples.
best_rf <- select_best(rf_tuned, metric = "rmse")
best_rf

final_wf <- finalize_workflow(corn_wf, best_rf)
final_wf
```



```{r }
#| message: false
#| warning: false
# ------------AGGREGATE TRAINING TRAITS-------------
train_trait_clean <- train_trait %>%
  mutate(
    date_planted = mdy(date_planted),
    date_harvested = mdy(date_harvested)
  ) %>%
  group_by(site, year, hybrid) %>%
  summarize(
    yield = mean(yield_mg_ha, na.rm = TRUE),
    grain_moisture = mean(grain_moisture, na.rm = TRUE),
    date_planted = min(date_planted, na.rm = TRUE),
    date_harvested = max(date_harvested, na.rm = TRUE),
    .groups = "drop"
  )
```


# Load packages

```{r}
#| message: false
#| message: false

library(tidymodels)
library(tidyverse)
library(recipes)  # step_rm, step_impute_* etc.
library(lubridate)
library(ranger)
library(rsample)
```

# Loading and inspecting the training and testing files

```{r}
#| message: false
#| message: false

# list.files("../data/training", full.names = TRUE)
# list.files("../data/testing", full.names = TRUE)

```

## **Read all 6 CSVs**

```{r}
#| message: false
#| message: false

train_meta  = read_csv("../data/training/training_meta.csv")
train_soil  = read_csv("../data/training/training_soil.csv")
train_trait = read_csv("../data/training/training_trait.csv")

test_meta   = read_csv("../data/testing/testing_meta.csv")
test_soil   = read_csv("../data/testing/testing_soil.csv")
test_sub    = read_csv("../data/testing/testing_submission.csv")
```

# Inspect column names

```{r}
#| message: false
#| message: false

# names(train_meta)
# names(train_soil)
# names(train_trait)
# 
# names(test_meta)
# names(test_soil)
# names(test_sub)
```

## **Join meta and soil into trait/submission**

```{r}
#| message: false
#| message: false

# Combine META and SOIL for training and testing data
train_env = train_meta %>%
  left_join(train_soil, by = c("year", "site"))

test_env = test_meta %>%
  left_join(test_soil, by = c("year", "site"))

# Add env info to trait
train_df = train_trait %>%
  left_join(train_env, by = c("year", "site"))

test_df = test_sub %>%
  left_join(test_env, by = c("year", "site"))

# checks 
glimpse(train_df)
glimpse(test_df)

# EDA
summary(train_df$yield_mg_ha)
table(train_df$year)
table(train_df$site)
```

## **Site-year grouping variable for CV use**

```{r}
#| message: false
#| message: false

train_df = train_df %>%
  mutate(site_year = paste(site, year, sep = "_"))
train_df

test_df = test_df %>%
  mutate(site_year = paste(site, year, sep = "_"))
test_df
```

## **Tidymodels and set-up of recipe**

```{r}
#| message: false
#| message: false

# convert character predictors to factors
train_df <- train_df %>%
  mutate(
    site          = factor(site),
    hybrid        = factor(hybrid),
    previous_crop = factor(previous_crop),
    site_year     = factor(site_year)
  )

test_df <- test_df %>%
  mutate(
    site          = factor(site),
    hybrid        = factor(hybrid),
    previous_crop = factor(previous_crop),
    site_year     = factor(site_year)
  )

# date parsing helper funvtion (month/day/year to Date)
parse_mdy_year <- function(x, year) {
  # x is like "5/5/14"; we know the numeric year column separately
  # so we coerce using lubridate::mdy with paste
  suppressWarnings(
    lubridate::mdy(paste(x, year))
  )
}

# count of predictor columns BEFORE recipe steps
n_pred <- train_df %>%
  select(-yield_mg_ha, -site_year) %>%
  ncol()

n_pred

```

# Define a recipe using yield_mg_ha as outcome

```{r}
#| message: false
#| message: false

corn_rec <- recipe(
  yield_mg_ha ~ year + site + replicate + block + hybrid +
    date_planted + date_harvested +
    grain_moisture +
    previous_crop + longitude + latitude +
    soilpH + om_pct + soilk_ppm + soilp_ppm,
  data = train_df
) %>%
  # Convert character predictors to factors
  step_string2factor(all_nominal_predictors()) %>%
  
  # Create Date variables from the character date columns
  step_mutate(
    planted_date   = mdy(date_planted),
    harvested_date = mdy(date_harvested)
  ) %>%
  
  # Engineer numeric date features
  step_mutate(
    doy_planted      = yday(planted_date),
    doy_harvested    = yday(harvested_date),
    harvest_interval = as.numeric(harvested_date - planted_date)
  ) %>%
  
  # Remove original string and intermediate Date columns
  step_rm(date_planted, date_harvested, planted_date, harvested_date) %>%
  
  # Impute missing numeric and categorical data
  step_impute_median(all_numeric_predictors()) %>%   # numeric median imputation[web:37][web:43]
  step_impute_mode(all_nominal_predictors()) %>%     # mode imputation for factors[web:40]
  
  # Dummy encode categorical predictors
  step_dummy(all_nominal_predictors())

corn_rec

```

## **Create grouped resampling**

```{r}
#| message: false
#| message: false

set.seed(123)

cv_folds <- group_vfold_cv(
  train_df,
  group = site_year,
  v = 5
)

corn_rec
cv_folds

```

## **Tuned XGBoost model with grouped CV**

```{r}
#| message: false
#| message: false

# XGBoost spec with tunable hyperparameters
xgb_spec <- boost_tree(
  trees          = 1000,
  tree_depth     = tune(),
  learn_rate     = tune(),
  loss_reduction = tune(),
  min_n          = tune(),
  sample_size    = tune(),
  mtry           = tune()
) %>%
  set_mode("regression") %>%
  set_engine("xgboost")

xgb_spec
```

# XGBoost workflow

```{r}
#| message: false
#| message: false

xgb_wf <- workflow() %>%
  add_model(xgb_spec) %>%
  add_recipe(corn_rec)

xgb_wf
```

## **Tuning grid**

```{r}
#| message: false
#| message: false

set.seed(123)

# Use a Latin space‑filling over a sensible range.
xgb_grid <- grid_space_filling(
  tree_depth(),
  learn_rate(range = c(-3, -0.5)),  # on log10 scale: 10^-3 to ~10^-0.5
  loss_reduction(),
  min_n(),
  sample_size = sample_prop(),
  mtry(range = c(5, min(50, n_pred))),  # use the n_pred counted from train_df
  size = 30
)

xgb_grid
```

# Tune with grouped CV

```{r}
set.seed(123)

xgb_tuned <- tune_grid(
  xgb_wf,
  resamples = cv_folds,
  grid      = xgb_grid,
  metrics   = metric_set(rmse, rsq),
  control   = control_grid(save_pred = TRUE)
)

xgb_tuned

collect_metrics(xgb_tuned) %>%
  filter(.metric == "rmse") %>%
  arrange(mean)

```

# XGB best parameters

```{r}
best_xgb <- select_best(xgb_tuned, metric = "rmse")
best_xgb

# Finalize the workflow
final_xgb_wf <- finalize_workflow(xgb_wf, best_xgb)
```

# Fit on all training data and predict 2024

```{r}
final_xgb_fit <- fit(final_xgb_wf, data = train_df)

xgb_preds <- predict(final_xgb_fit, new_data = test_df) %>%
  bind_cols(test_df %>% select(year, site, hybrid)) %>%
  rename(pred_yield_mg_ha = .pred)

xgb_preds
```




## **Specify a random forest model**

```{r}
# rf_spec <- rand_forest(
#   mtry  = tune(),   # number of predictors tried at each split (we'll tune)
#   trees = 500,      # reasonable default number of trees
#   min_n = tune()    # minimum data points in a terminal node
# ) %>%
#   set_mode("regression") %>%
#   set_engine("ranger", importance = "impurity")


rf_spec <- rand_forest(
  mtry  = 20,     # heuristic; can adjust later
  trees = 500,
  min_n = 5
) %>%
  set_mode("regression") %>%
  set_engine("ranger", importance = "impurity")
```

## **Build a workflow (recipe + model)**

```{r}
corn_wf <- workflow() %>%
  add_model(rf_spec) %>%
  add_recipe(corn_rec)

corn_wf
```

## **Evaluate with grouped CV (no tuning)**

```{r}
set.seed(123)

rf_res <- fit_resamples(
  corn_wf,
  resamples = cv_folds,
  metrics   = metric_set(rmse, rsq),
  control   = control_resamples(save_pred = TRUE)
)

collect_metrics(rf_res)
```

## **Fit final model on all training data and predict on test**

```{r}
final_fit <- fit(corn_wf, data = train_df)
```

# Generate predictions for 2024

```{r}
test_pred <- predict(final_fit, new_data = test_df) %>%
  bind_cols(test_df %>% select(year, site, hybrid))

# Rename prediction column
test_pred <- test_pred %>%
  rename(pred_yield_mg_ha = .pred)

test_pred

write_csv(test_pred, "../output/corn_yield_predictions_rf.csv")
```

## **Set up tuning grid and metrics**

```{r}
# metrics
corn_metrics <- metric_set(rmse, rsq)  # yardstick metrics for numeric outcomes[web:93][web:90]

# a small tuning grid
rf_grid <- grid_regular(
  mtry(range = c(5L, 50L)),   # you can widen later
  min_n(range = c(5L, 25L)),
  levels = 5                  # 5x5 = 25 combos
)
```

## **Tune with grouped CV**

```{r}
# training step
set.seed(123)

rf_tuned <- tune_grid(
  corn_wf,
  resamples = cv_folds,
  grid      = rf_grid,
  metrics   = corn_metrics,
  control   = control_grid(save_pred = TRUE)
)

rf_tuned
```

# inspect results

```{r}
collect_metrics(rf_tuned) %>%
  arrange(mean)
```

## **best parameters and finalize**

```{r}
# picks the parameter combination with lowest RMSE over the resamples.
best_rf <- select_best(rf_tuned, metric = "rmse")
best_rf

final_wf <- finalize_workflow(corn_wf, best_rf)
final_wf
```
